---
title: "Quantitative Genetics: Diallel Analysis with sommer"
author: "Bioinformatics Portfolio"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    theme: united
---

# Introduction

This analysis evaluates a half-diallel mating design using the `sommer` package in R. The goal is to partition genetic variance into **General Combining Ability (GCA)** and **Specific Combining Ability (SCA)** and to estimate breeding values (BLUPs) for parent selection.

# 1. Environment Setup

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

# Check and install necessary packages
packages <- c("sommer", "dplyr", "ggplot2")
new_packages <- packages[!(packages %in% installed.packages()[,"Package"])]
if(length(new_packages)) install.packages(new_packages)

library(sommer)
library(dplyr)
library(ggplot2)
```

# 2. Data Exploration

The `DT_halfdiallel` dataset contains sugar content measurements from a sugar beet half-diallel cross.

```{r}
data("DT_halfdiallel")
DT <- DT_halfdiallel
head(DT)

# Summary of the experimental design
cat("Number of Parents:", length(unique(c(as.character(DT$female), as.character(DT$male)))), "\n")
cat("Number of Crosses:", length(unique(DT$geno)), "\n")
cat("Number of Replicates:", length(unique(DT$block)), "\n")

# Visualize raw cross means
cross_means <- DT %>%
  group_by(geno) %>%
  summarise(meanSugar = mean(sugar, na.rm = TRUE)) %>%
  arrange(desc(meanSugar))

ggplot(cross_means, aes(x = reorder(geno, meanSugar), y = meanSugar)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Raw Mean Sugar Content per Cross",
       x = "Cross (Geno)", y = "Sugar Content") +
  theme_minimal()
```

# 3. Mixed Model Fitting (GCA + SCA)

We fit a model where the parent effects (GCA) are combined using an overlay matrix, and specific combinations (SCA) are treated as random effects.

```{r}
# Prepare factors
DT$femalef <- as.factor(DT$female)
DT$malef <- as.factor(DT$male)
DT$genof <- as.factor(DT$geno)

# Fit the model
# GCA: overlay(femalef, malef)
# SCA: genof
mod1 <- mmes(sugar ~ block,
             random = ~ vsm(ism(overlay(femalef, malef))) + vsm(ism(genof)),
             rcov = ~ units,
             data = DT)

summary(mod1)
```

# 4. Heritability Estimation

We use `vpredict` to calculate narrow-sense ($h^2$) and broad-sense heritability ($H^2$).

```{r}
# Variance components
vars <- summary(mod1)$varcomp

# Narrow-sense heritability (Additive / Total)
h2_res <- vpredict(mod1, h2 ~ V1 / (V1 + V2 + V3))
cat("Narrow-sense heritability (h²):", round(h2_res$Estimate, 3), "\n")

# Broad-sense heritability ((Additive + Dominance/SCA) / Total)
H2_res <- vpredict(mod1, H2 ~ (V1 + V2) / (V1 + V2 + V3))
cat("Broad-sense heritability (H²):", round(H2_res$Estimate, 3), "\n")
```

# 5. BLUPs of Genetic Effects

Best Linear Unbiased Predictors (BLUPs) provide a more accurate estimate of genetic merit than raw means by shrinking estimates toward the mean based on the reliability of the data.

```{r}
# Extract GCA BLUPs (Parents)
gca <- mod1$uList[[1]]
gca_df <- data.frame(
  Parent = levels(DT$femalef),
  GCA_BLUP = as.numeric(gca)
) %>% arrange(desc(GCA_BLUP))

print("Top Parents by GCA:")
head(gca_df)

# Extract SCA BLUPs (Crosses)
sca <- mod1$uList[[2]]
sca_df <- data.frame(
  Cross = levels(DT$genof),
  SCA_BLUP = as.numeric(sca)
) %>% arrange(desc(SCA_BLUP))

print("Top Crosses by SCA:")
head(sca_df)
```

# 6. Comparing Raw Means vs. BLUPs

```{r}
# Merge raw means with SCA BLUPs
comparison <- inner_join(cross_means, sca_df, by = c("geno" = "Cross"))

# Correlation
correlation <- cor(comparison$meanSugar, comparison$SCA_BLUP)

# Plot
ggplot(comparison, aes(x = meanSugar, y = SCA_BLUP)) +
  geom_point(color = "darkred") +
  geom_smooth(method = "lm", se = FALSE, color = "blue", linetype = "dashed") +
  labs(x = "Raw Means", y = "SCA BLUPs",
       title = paste0("Shrinkage Effect: Raw Means vs. BLUPs (r = ", round(correlation, 2), ")")) +
  theme_minimal()
```

# Conclusion

The analysis demonstrates that parent selection based on GCA BLUPs is more robust than simple phenotypic selection. The high narrow-sense heritability suggests that additive genetic effects are substantial, indicating that progeny performance is predictable from parental GCA.
